Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-10-18 WholePlotTrajectories.Rmd
Jecol review

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Hérault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: >   
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the controversy questions the extent of communities’ resilience regarding their taxonomic and functional facets. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed the tree community trajectories over 30 years following a disturbance gradient in a Neotropical forest. Specifically, we examined community functional and taxonomic trajectories with regards to diversity, composition and redundancy. Functional trajectories were drawn based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict communities taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Determinants of Plant Community Diversity and Structure, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::word_document2: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
---

@@

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes increases the pressure on remaining natural forests [@Morales-Hidalgo2015] and threatens the maintenance and dynamics in space and time of communities structure, composition and functioning [@Anderson-Teixeira2013].

In tropical forests, ecological communities are regularly re-shaped by natural disturbance events changing both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions such as competition among species [@Chesson2000]. 
One of the cornerstone of tropical forest ecology is to understand the processes and drivers of ecosystems response to disturbance [@Chazdon2003a]. 
For now, this has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016] that are rapid and convenient to measure. 
These structural parameters have been successfully modeled, giving important insights into the recovery of ecosystem processes and services [@Herault2018]. 
However the response of forests diversity and composition remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014; @Liang2016].
In the short-term, moderate disturbance may lead to positive impacts on communities diversity, an idea formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity when disturbance intensity is not too high [@Molino2001;@Kariuki2006a].

Validations of the IDH though remain scarce in the long-term and mainly rely on the analysis of taxonomic richness [@Molino2001]. Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. More ecological-meaningful analysis would couple richness with (i) evenness, that would reveal the changes in the species abundance distribution and thus the underlying ecological processes, and (ii) composition that is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. Furthermore, a functional approach accounting for species biological attributes would directly link communities diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was successfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually a complete overview of community response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. 
We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. 
Specifically, we (i) elucidated community taxonomic and functional recovery and the underlying ecological processes, (ii) clarified the validity of the IDH in the long term for tropical forest and its translation into different trajectories in time, and (iii) questioned community recovery time.

@@

Paracou station in French Guiana (5\textdegree 18'N and 52\textdegree 53'W) is located in a lowland tropical rain forest in a tropical wet climate with mean annual temperature of 26\textdegree C, mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) and a 3-month dry season (< 100 mm.month^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. 
Elevation ranges between 5 and 50 m and soils correspond to thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains.

The experiment is a network of twelve 6.25ha plots that underwent a disturbance gradient of three logging, thinning and fuelwood cutting treatments (Table \@ref(tab:Tab1)) according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Schmitt1990]. 
The disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance [@Piponiot2016] estimated with the BIOMASS R package [@Biomass2018].

@@

The study site corresponds to a tropical rainforest typical of the Guiana Shield with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae. 
In the twelve experimental plots, all trees above 10 cm DBH have been mapped and measured annually since 1984. 
Trees are first identified with a vernacular name assigned by the forest worker team, and afterward with a scientific name assigned by botanists during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas more infrequent ones were identified under general identifiers only distinguishing trees and palm trees. 
From 2003, botanical campaigns have been conducted every 5 to 6 years to identify all trees at the species level but identification levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as vernacular names usually correspond to different botanical species. 
It resulted in significant taxonomic uncertainties that had to be propagated to composition and diversity metrics.
The uncertainty propagation was done through a Bayesian framework reconstituting complete inventories at genus level from real incomplete ones on the basis of vernacular/botanical names association. 
Vernacular names were replaced through multinomial trials based on the association probability $\big[\alpha_1, \alpha_2,..., \alpha_V\big]$ observed across all inventories between each vernacular name *v* and all species $\big[s_1, s_2,..., s_N\big]$:

@@

Six functional traits representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area) and stem economics (wood specific gravity and bark thickness), and life-history traits (maximum specific height and seed mass) came from the BRIDGE project [^1]. Trait values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou, and comprised 294 species pertaining to 157 genera. 
Missing trait values (10%) were filled using multivariate imputation by chained equation [@Mice2011].
Imputations were restricted within genus or family when samples were too scarce, in order to account for the phylogenetic signal. 
Whenever a species was not in the dataset, it was attributed a set of trait values randomly sampled among species of the next higher taxonomic level (same genus or family). 
As seed mass information was classified into classes, no data filling process was applied and analyses were restricted to the 414 botanical species recorded.

@@

To counter taxonomic uncertainties due to the variability of botanical identification levels (in space) and protocols (in time), the taxonomic composition and diversity analysis were conducted at the genus level. 
Taxonomic and functional trajectories of community composition were followed in a two-dimensional NMDS ordination space.
Two NMDS using abundance-based (Bray-Curtis) dissimilarity measures were conducted to map either taxonomic or functional composition, the later based on the 7 leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the Euclidean distance between the target inventories and the reference inventories in 1989, *i.e.* 2 years after disturbance. 
Univariate trajectories of the leaf, stem and life-history traits were also visualized with the community weighted means (CWM) [@Diaz2007]. 
Species seed mass corresponded to 5 classes of increasing mass, seed mass trajectories were therefore reported as the proportion of each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and evenness, *i.e* the Hill number translation of the Simpson index [@Hill1973].
These indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0 and 2 order of diversity (q), recommended for diversity studies [@Marcon2015b]. 
The functional diversity was reported using the functional richness and functional evenness, *i.e* Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the spearman rank correlation between the extrema of taxonomic and functional metrics reached over the 30 years and the initial %AGB loss. 
They were besides analyzed through polynomial regression between (i) taxonomic and functional richness and evenness and (ii) the initial %AGB loss at 10, 20 and 30 years after disturbance.

The functional redundancy was measured as the overlap among species in community functional space [@Carmona2016]. 
The samples of the trait database were first mapped in a 2-dimensional plan with a PCA analysis. 
Then, multivariate kernel density estimator associated with individual trees returned species traits probability distribution (TDP).
Species TDP weighted by species abundance were eventually summed for each community.
Community functional redundancy was the sum of TDPs overlap, expressed as the average number of species that could be removed from without reducing the functional space (Supp. Mat. - Fig. S1 for a more comprehensive scheme).

@@

## Communities Composition

From 1989 (2 years after disturbance) to 2015 (28 years after disturbance), 828 388 individual trees and 591 botanical species pertaining to 223 genus and 64 families were recorded.

While both taxonomic and functional composition remained stable in undisturbed communities (Fig. \@ref(fig:NMDSplans)), they followed marked and consistent trajectories over post-\break disturbance time.
In disturbed communities, these compositional changes corresponded to shifts towards species with more acquisitive functional strategies, from communities with high average WSG to high average SLA and chlorophyll content (see appendix I). For functional composition, this translated into cyclic compositional changes with an unachieved recovery of the initial composition (Fig. \@ref(fig:NMDSplans)). The maximum dissimilarity with the initial state was positively correlated to the disturbance intensity for both taxonomic and functional composition ($\rho_{spearman}^{Taxonomic}=0.87$ and $\rho_{spearman}^{Functional}=0.90$ respectively). The maximum value was reached around 26 years after disturbance for taxonomic composition and 22 years for functional composition.

```{r NMDSplans, out.width = '100%', echo=FALSE, fig.cap="Plot trajectories in terms of flora composition (left panels (ref:Bolda) and (ref:Boldc)) and functional composition (right panels (ref:Boldb) and (ref:Boldd)) in a two-dimensional NMDS space. Lower panels ((ref:Boldc) and (ref:Boldd)) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.", fig.env="figure*"}
load("DB/FunctionalComposition_forGraphs");load("DB/TaxoComposition_forGraphs")
load("DB/FunDistance_ForGraphs");load("DB/TaxoDistance_ForGraphs")

@@

Except for leaf chlorophyll content, which continued to increase for some T3 and T2 plots 30 years after disturbance, all traits and seed mass proportions followed unimodal trajectories either stabilizing or returning towards their initial values.

Maximum height at adult stage (*Hmax*), leaf toughness and wood specific gravity (*WSG*) first decreased and then slightly increased but remained significantly lower than their initial value (Fig. \@ref(fig:CWM)). On the other side, bark thickness and specific leaf area (*SLA*) increased and while bark thickness remained substantially high after 30 years, *SLA* had almost recovered its initial value. For all traits, the maximum difference to initial value was correlated to the disturbance intensity ($\rho_{spearman}^{Leaf thickness}=0.76$, $\rho_{spearman}^{Chlorophyll content}=0.60$, $\rho_{spearman}^{Leaf toughness}=-0.53$, $\rho_{spearman}^{SLA}=0.93$, $\rho_{spearman}^{WSG}=-0.75$, $\rho_{spearman}^{Bark thickness}=0.71$, $\rho_{spearman}^{Hmax}=-0.40$).
The proportions of the three lightest seed mass classes increased in all disturbed plots, and decreased after 30 years for the lightest class while it stabilized for the two other (Supp. Mat. - Fig. S2).

```{r CWM, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the communities weighted means over 30 years after disturbance of four leaf traits (Leaf thickness, chlorophyll content, toughness, and specific area), two stem traits (wood specific gravity, and bark thickness) and one life history trait (Specific maximum height at adult stage). ", fig.env="figure*"}
load("DB/CWM")

@@

## Communities richness and evenness

For undisturbed plots, taxonomic Richness and Evenness remained stable over the 30 years monitored.
In disturbed communities, after low disturbance intensity the taxonomic richness increased, reaching a maximum gain of 14 botanical genera (plot 3 from treatment 2). After intense disturbance the taxonomic richness followed a more complex trajectory, decreasing for ten years after disturbance before recovering to pre-disturbance values. The maximum richness loss or gain after disturbance was positively correlated to the disturbance intensity ($\rho_{spearman}^{Richness}=0.50$).
In all disturbed plots the taxonomic evenness first increased until a maximum reached after around 20 years. This maximum was positively correlated to the disturbance intensity ($\rho_{spearman}^{Evenness}=0.77$).
The evenness then stabilized except for two T3 plots (plots 8 and 12) for which evenness kept increasing.

```{r DivTaxo, out.width = '100%', echo=FALSE, fig.cap="Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic (ref:Bolda) richness, (ref:Boldb), taxonomic evenness, (ref:Boldc) functional richness, and (ref:Boldd) functional evenness. Shaded areas are the credibility intervals ", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")
load("DB/FunctionalTraj_ForGraphs")
load("DB/FunctionalRichnessTraj_ForGraphs")

@@

The plot 7 from treatment 1 displayed constantly outlying functional richness and evenness and was removed from the graphical representation for better readability.
In undisturbed plots both functional richness and evenness remained stable along the 30 years. 
In disturbed plots, functional richness and evenness trajectories depended on the disturbance intensity with their maximum positively correlated to %AGB loss $\rho_{spearman}^{Richness}=0.76$ and $\rho_{spearman}^{Evenness}=0.60$.
Functional richness and evenness displayed for low disturbance intensity a low but long-lasting increase up to a maximum reached after 20-25 years, and for high intensity, a fast but short increase followed after 10 years by a slow decrease towards the initial values.

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the controversy questions the extent of community resilience regarding their taxonomic and functional facets. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed the tree community trajectories over 30 years following a disturbance gradient in a Neotropical forest. Specifically, we examined community functional and taxonomic trajectories with regards to diversity, composition and redundancy. Functional trajectories were drawn based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict community taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::word_document2: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
---

@@

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes increases the pressure on remaining natural forests [@Morales-Hidalgo2015] and threatens the maintenance and dynamics in space and time of community structure, composition and functioning [@Anderson-Teixeira2013].

In tropical forests, ecological communities are regularly re-shaped by natural disturbance events changing both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions such as competition among species [@Chesson2000]. 
One of the cornerstone of tropical forest ecology is to understand the processes and drivers of ecosystems response to disturbance [@Chazdon2003a]. 
For now, this has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016] that are rapid and convenient to measure. 
These structural parameters have been successfully modeled, giving important insights into the recovery of ecosystem processes and services [@Herault2018]. 
However the response of forests diversity and composition remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014; @Liang2016].
In the short-term, moderate disturbance may lead to positive impacts on community diversity, an idea formalized by the intermediate disturbance hypothesis (IDH) which predicts a maximum species diversity when disturbance intensity is not too high [@Molino2001;@Kariuki2006a].

Validations of the IDH though remain scarce in the long-term and mainly rely on the analysis of taxonomic richness [@Molino2001]. Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. More ecological-meaningful analysis would couple richness with (i) evenness, which would reveal the changes in the species abundance distribution and thus the underlying ecological processes, and (ii) composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. Furthermore, a functional approach accounting for species biological attributes would directly link community diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was successfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually, a complete overview of community response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. 
We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. 
Specifically, we (i) elucidated community taxonomic and functional recovery and the underlying ecological processes, (ii) clarified the validity of the IDH in the long term for tropical forest and its translation into different trajectories in time, and (iii) examined community recovery time.

@@

Paracou station in French Guiana (5\textdegree 18'N and 52\textdegree 53'W) is located in a lowland tropical rain forest in a tropical wet climate with mean annual temperature of 26\textdegree C, mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) and a 3-month dry season (< 100 mm.month^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. 
Elevation ranges between 5 and 50 m and soils correspond to thin acrisols over a layer of transformed saprolite with low permeability which produces lateral drainage during heavy rains.

The experiment is a network of 12 6.25 ha plots that underwent a disturbance gradient of three logging, thinning and fuelwood cutting treatments (Table \@ref(tab:Tab1)) according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to an intensity gradient. For treatment 1 (T1) 10 trees of commercial species (of a diameter at 1.3 m height (DBH) equal or above 50 cm) were felled per hectare. For treatment 2 (T2) 10 trees/ha of commercial species (DBH  $\geq$ 50 cm) were felled and 30 trees/ha of non-valuable species (DBH $\geq$ 40 cm) were removed by poison girdling. For treatment 3 (T3) 10 trees/ha of commercial species (DBH $\geq$ 50 cm) were felled and 30 trees/ha of non-valuable species (15 with DBH $\geq$ 50 cm and 15 with DBH $\geq$ 40 cm) were removed by poison girdling [@Schmitt1990]. 
Disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance [@Piponiot2016] estimated with the BIOMASS R package [@Biomass2018].

@@

The study site corresponds to a tropical rainforest typical of the Guiana Shield with a dominance of *Fabaceae*, *Chrysobalanaceae*, *Lecythidaceae* and *Sapotaceae*. 
In the 12 experimental plots, all trees above 10 cm DBH have been mapped and measured annually since 1984. 
Trees are first identified with a vernacular name assigned by the forest worker team, and afterward with a scientific name assigned by botanists during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas more infrequent ones were identified under general identifiers only distinguishing trees and palms. 
From 2003, botanical campaigns have been conducted every 5 to 6 years to identify all trees at the species level but identification levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as vernacular names usually correspond to different botanical species. 
This resulted in significant taxonomic uncertainty that had to be propagated to composition and diversity metrics.
The uncertainty propagation was done through a Bayesian framework reconstituting complete inventories at genus level from real incomplete ones on the basis of vernacular/botanical names association. 
Vernacular names were replaced through multinomial trials based on the association probability $\big[\alpha_1, \alpha_2,..., \alpha_V\big]$ observed across all inventories between each vernacular name *v* and all species $\big[s_1, s_2,..., s_N\big]$:

@@

Six functional traits representing leaf economics (leaf thickness, toughness, total chlorophyll content and specific leaf area) and stem economics (wood specific gravity and bark thickness), and life-history traits (maximum specific height and seed mass) came from the BRIDGE project [^1]. Trait values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou, and comprised 294 species pertaining to 157 genera. 
Missing trait values (10%) were filled using multivariate imputation by chained equation [@Mice2011].
Imputations were restricted within genus or family when samples were too scarce, in order to account for the phylogenetic signal. 
Whenever a species was not in the dataset, it was attributed a set of trait values randomly sampled among species of the next higher taxonomic level (same genus or family). 
As seed mass information was classified into classes, no data filling process was applied and analyses were restricted to the 414 botanical species recorded.

@@

Because of the variability in the precision of botanical identification efforts, we were constraint to conduct the taxonomic composition and diversity analysis at the genus level. 
Taxonomic and functional trajectories of community composition were followed in a two-dimensional NMDS ordination space.
Two NMDS using abundance-based (Bray-Curtis) dissimilarity measures were conducted to map either taxonomic or functional composition, the latter based on the seven leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the Euclidean distance between the target inventories and the reference inventories in 1989, *i.e.* 2 years after disturbance. 
Univariate trajectories of the leaf, stem and life-history traits were also visualized with the community weighted means (CWM) [@Diaz2007]. 
Species seed mass were given in 5 mass classes.
Seed mass trajectories were therefore reported as the proportion of each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the Hill number translation of the Simpson index [@Hill1973].
The comparison between richness and Simpson diversity assessed the variations in taxonomic evenness. 
Both indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0 and 2 order of diversity (q), recommended for diversity studies [@Marcon2015b]. 
The functional diversity was reported using the functional richness and functional evenness, *i.e* the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the Spearman rank correlation between the extremes of taxonomic and functional metrics reached over the 30 years and the initial %AGB loss. 
They were besides analyzed through polynomial regression between (i) taxonomic and functional richness and evenness and (ii) the initial %AGB loss at 10, 20 and 30 years after disturbance.

Functional redundancy was measured as the overlap among species in community functional space [@Carmona2016]. 
First, the individuals of the trait database were mapped in the plane of the two first axes from a PCA analysis. Then, for each species, the traits probability density (TPD) were computed through multivariate kernel density estimators. Second, for each community, the functional traits probability density were computed as the sum of the TDP of the inventoried species weighted by species abundance. Third, the whole functional space was divided into a 100 x 100 grid, before counting for each cell the number of species which TPD is positive. The average count across cells, minus 1, gives the average number of species in the community that share the same traits, which is the Community Functional Redundancy.

@@

## Community Composition

From 1989 (2 years after disturbance) to 2015 (28 years after disturbance), 828 388 individual trees and 591 botanical species pertaining to 223 genera and 64 families were recorded.

While both taxonomic and functional composition remained stable in undisturbed communities (Fig. \@ref(fig:NMDSplans)), they followed marked and consistent trajectories over post-\break disturbance time.
In disturbed communities, these compositional changes corresponded to shifts towards species with more acquisitive functional strategies, from communities with high mean WSG to high mean SLA and chlorophyll content (see appendix I). 
For functional composition, this translated into cyclic compositional changes with an incomplete recovery of the initial composition (Fig. \@ref(fig:NMDSplans)). The maximum dissimilarity with the initial state was positively correlated with the disturbance intensity for both taxonomic and functional composition ($\rho_{Spearman}^{Taxonomic}=0.87$ and $\rho_{Spearman}^{Functional}=0.90$ respectively). 
The maximum dissimilarity with the initial was reached for taxonomic composition between 15 to 25 years, for most of the plots, and around 22 years for functional composition.

```{r NMDSplans, out.width = '100%', echo=FALSE, fig.cap="Plot trajectories in terms of taxonomic composition ((ref:Bolda) and (ref:Boldc)) and functional composition ((ref:Boldb) and (ref:Boldd)) in a two-dimensional NMDS space. Lower panels ((ref:Boldc) and (ref:Boldd)) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.", fig.env="figure*"}
load("DB/FunctionalComposition_forGraphs");load("DB/TaxoComposition_forGraphs")
load("DB/FunDistance_ForGraphs");load("DB/TaxoDistance_ForGraphs")

@@

Community CWM average value of all traits and seed mass proportions followed unimodal trajectories, either stabilizing or returning towards their initial values, to the exception of leaf chlorophyll content, which continued to increase for some T3 and T2 plots 30 years after disturbance.

Community CWM average value of Maximum height at adult stage (*Hmax*), leaf toughness and wood specific gravity (*WSG*) first decreased and then slightly increased but remained significantly lower than their initial value (Fig. \@ref(fig:CWM)). On the other side, bark thickness and specific leaf area (*SLA*) increased and while bark thickness remained substantially high after 30 years, *SLA* had almost recovered to its initial value. For all traits, the maximum difference to initial value was correlated to the disturbance intensity.
Positive correlations were observed for Leaf thickness, chlorophyll content, SLA and bark thickness ($\rho_{Spearman}^{Leaf thickness}=0.76$, $\rho_{Spearman}^{Chlorophyll content}=0.60$, $\rho_{Spearman}^{SLA}=0.93$, $\rho_{Spearman}^{Bark thickness}=0.71$).
Negative correlation was observed for Leaf toughness, WSG and Hmax ($\rho_{Spearman}^{Leaf toughness}=-0.53$, $\rho_{Spearman}^{WSG}=-0.75$, $\rho_{Spearman}^{Hmax}=-0.40$)
The proportions of the three lightest seed mass classes increased in all disturbed plots, and decreased after 30 years for the lightest class while it stabilized for the two other (Supp. Mat. - Fig. S2).

```{r CWM, out.width = '100%', echo=FALSE, fig.cap="Trajectories of community weighted means over 30 years after disturbance of four leaf traits (Leaf thickness, chlorophyll content, toughness, and specific area), two stem traits (wood specific gravity, and bark thickness) and one life history trait (Specific maximum height at adult stage). ", fig.env="figure*"}
load("DB/CWM")

@@

## Community taxonomic and functional diversity

For undisturbed plots, taxonomic richness and Simpson diversity remained stable over the 30 years of monitoring.
In disturbed communities, after low disturbance intensity the taxonomic richness increased, reaching a maximum gain of 14 botanical genera (plot 3 from treatment 2). After intense disturbance the taxonomic richness followed a more complex trajectory, decreasing for ten years after disturbance before recovering to pre-disturbance values. The maximum richness loss or gain after disturbance was positively correlated with the disturbance intensity ($\rho_{Spearman}^{Richness}=0.50$).
In all disturbed plots the Simpson diversity first increased until a maximum reached after around 20 years. This maximum was positively correlated with the disturbance intensity ($\rho_{Spearman}^{Simpson}=0.77$).
The Simpson diversity then stabilized except for two T3 plots (plots 8 and 12) for which it kept increasing (Fig. \@ref(fig:DivTaxo)).

```{r DivTaxo, out.width = '100%', echo=FALSE, fig.cap="Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic richness (ref:Bolda), Simpson diversity (ref:Boldb), functional richness (ref:Boldc), and Rao diversity (ref:Boldd). Shaded areas are the credibility intervals ", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")
load("DB/FunctionalTraj_ForGraphs")
load("DB/FunctionalRichnessTraj_ForGraphs")

@@

The plot 7 from treatment 1 displayed constantly outlying functional richness and Rao diversity and was removed from the graphical representation for better readability.
In undisturbed plots both functional richness and Rao diversity remained stable along the 30 years. 
In disturbed plots, both trajectories depended on the disturbance intensity, with their maximum being positively correlated to %AGB loss $\rho_{Spearman}^{Richness}=0.76$ and $\rho_{Spearman}^{Rao}=0.60$.
Functional richness and Rao diversity displayed for low disturbance intensity a low but long-lasting increase up to a maximum reached after 20-25 years, and for high intensity, a fast but short increase followed after 10 years by a slow decrease towards the initial values.

The second-degree polynomial regressions between (i) the percentage AGB loss and (ii) taxonomic and functional diversity after 10, 20 and 30 years best predicted the hump-shaped curve of the disturbance impact along the disturbance intensity gradient \@ref(fig:IDHplot).
The relationship between the disturbance impact and its intensity was more markedly hump-shaped for the taxonomic richness than for the Simpson diversity. For both functional richness and Rao diversity the relationship was almost linear.
The regression model better predicted the functional richness and Rao diversity ($0.55<R^2_{Functional Richness}<0.72$, and $0.60<R^2_{Functional Rao}<0.81$) than the taxonomic richness and evenness ($0.21<R^2_{Taxonomic Richness}<0.4$, and $-0.15<R^2_{Taxonomic Simpson}<0.43$ respectively)
